\documentclass [PhD] {uclathes}
\usepackage{graphicx}
\usepackage[table]{xcolor}
\usepackage{epsfig}
\usepackage{cite}
\usepackage{verbatim}
\usepackage{subfigure}
\usepackage{caption}
\usepackage{listings}
\usepackage{color}
\lstset{language=Ruby, basicstyle=\footnotesize, frame=single, showspaces=false, showstringspaces=false}

% \input {mymacros}                         % personal LaTeX macros

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Usually things live in separate flies.
%
% \input {prelim}                           % preliminary page info

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                      %
%                          PRELIMINARY PAGES                           %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title          {Communication Paradigms for Mobile Ad Hoc Networks}
\author         {Justin Scott Collins}
% Note: department is really your area of research.  I.e. leave out 'Department of'.
\department     {Computer Science}
% Note:  degreeyear should be optional, but as of  5-Feb-96
% it seems required or you get a year of ``2''.   -johnh
\degreeyear     {2014}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chair          {Rajive Bagrodia}
\member         {Greg Pottie}
\member         {Stott Parker}
\member         {Mario Gerla}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\dedication     {
To my mom, without whom I wouldn't be.

To my wife, without whom I couldn't be.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\acknowledgments {
  I would like to thank my advisor Rajive Bagrodia for his patience, guidance, and belief that I would eventually complete this research.

  Sincere thanks to my managers at AT\&T Interactive and Twitter who were gracious enough to pay me, mentor me, and allow me the flexibility to complete my degree at the same time.

  Many thanks to the many who inspired me to study computer science. Special thanks to my uncle Michael Rudd who gave me his old TRS-80 Model 100 and permanently changed my career path from hardware to software and has remained a source of encouragement and academic advice. Thank you to David Joslin who provided my first research experience and publication.

  I am very grateful to have been a part of the Computer Science Graduate Student Committee (CSGSC). I would especially like to thank the founding members who were friendly enough to include me and made several years much more pleasant.

  I owe a surprising amount to Tom Murphy VII, who helped me become president of my high school, motivated me to investigate programming languages, led me to pursue a PhD, and inspired many of my projects. I hope to meet him some day.

  Finally, there is no way I would have made it this far without the advice and love from my mom, nor the incredible patience, understanding, and love of my wife. Thank you.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%
% UCLA Policy on Vita
% For security reasons, UCLA policy specifies that your birth year and birth place should not be included in your vita.
%
\vitaitem   {2002--2005}
                {Helpdesk Specialist,
                Seattle University Law School, Seattle, Washington.}
\vitaitem   {2005--2006}
                {Systems Administrator,
                Klir Technologies, Seattle, Washington.}
\vitaitem   {2006}
                {B.S.~(Computer Science),
                Seattle University.}
\vitaitem   {2007--2010}
                {Graduate Student Researcher, Computer Science Department, UCLA.}
\vitaitem   {2009}
                {M.S.~(Computer Science),
                UCLA.}
\vitaitem   {2010-2012}
                {Security Engineer, AT\&T Interactive, Glendale, California.}
\vitaitem   {2012--present}
                {Security Engineer, Twitter, Inc., San Francisco, California.}
\vitaitem   {2013--present}
                {Technology and Security Advisor, Guide Financial, San Francisco, California.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\publication    {J. Collins and D. Joslin. ``Improving genetic algorithm performance with intelligent mappings from chromosomes to solutions." In Proc. of the 8th Annual Conference on Genetic and Evolutionary Computation (GECCO '06) (Seattle, Washington, USA, July 08 - 12, 2006).}

\publication    {D. Joslin and J. Collins. ``Greedy transformation of evolutionary algorithm search spaces for scheduling problems." IEEE Congress on Evolutionary Computation, 2007 (CEC 2007).}

\publication    {Justin Collins and Rajive Bagrodia. ``Programming in Mobile Ad Hoc Networks." In Proceedings of the Fourth International Wireless Internet Conference (WICON 2008).}

\publication    {Justin Collins and Rajive Bagrodia. ``A Quantitative Comparison of Communication Paradigms for MANETs." In Proc. of the 7th International ICST Conference on Mobile and Ubiquitous Systems: Computing, Networking and Services (MobiQuitous 2010).}

\presentation	{2013 Fukuoka Ruby Award: Outstanding Performance. Presented in Fukuoka, Japan.}

\publication    {Justin Collins and Rajive Bagrodia. ``MELON: A Persistent Message-Based Communication Paradigm for MANETs." In Proc. of the 10th International ICST Conference on Mobile and Ubiquitous Systems: Computing, Networking and Services (MobiQuitous 2013).}

\publication    {Justin Collins and Rajive Bagrodia. ``MANET Application Development with MELON." In Proc. of the 13th International Conference on Ad-Hoc Networks and Wireless (ADHOC-NOW 2014).}

\presentation {``Mobile Ad Hoc Networks are Coming, But We Aren't Ready.'' Midwest.IO 2014, Kansas City, Missouri.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\abstract {
Consumer implementations of mobile ad hoc networks (MANETs) are rapidly becoming possible due to the proliferation and ubiquity of smartphones, tablets, and other portable computers. However, the combination of wireless communication, device mobility, and the self-organizing nature of MANETs presents a challenging environment for developing distributed applications. Networking middleware and libraries for MANET applications typically provide adaptations of traditional distributed computing paradigms designed for wired networks. In this work we quantitatively evaluate existing projects and their fundamental underlying communication paradigms using real applications in realistic MANET environments. We then propose and evaluate a new communication paradigm specifically designed for MANETs.
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%



\begin {document}
\makeintropages

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input {intro}
\input {communication_paradigms}
\input {analysis_of_existing}
\input {melon_model}
\input {melon_implementation}
\input {case_studies}
\input {evaluation}
\input {conclusion}
\input {appendix}

\bibliography {refs}    % bibliography references
\bibliographystyle {uclathes}

\end {document}

